Semantic Linearity in a Lazy Optimising Compiler

### 2025-07-10 icfp.tex
add reference to fractional permissions
Fix typos

The alternative judgement $\G;\D \vdash_{alt} \rho \to e :^z_\D \s \Rightarrow
\vp$ is used to type case alternatives, encompassing three
``modes'' that are distinguished by the kind of arrow that is used:
for alternatives of case expressions whose scrutinee is in WHNF
($\Rightarrow_\mathsf{WHNF}$), not in WHNF ($\Rightarrow_\mathsf{NWHNF}$),
and for alternatives agnostic to the WHNF-ness of the scrutinee
($\Rightarrow$), with $\Rightarrow$ also generalizing the other two.
%
% Following the $Case_\textrm{WHNF}$ rule in which we use the $\Mapsto$ alternative
% judgement, t

First, an alternative whose pattern is a constructor with $n > 0$ linear components
under the $\Rightarrow_{\mathsf{NWHNF}}$ mode is typed as:
\[
\TypeAltNWHNF
\]
The rule states that, for such a match on a scrutinee already in
WHNF, we introduce the linear components of the pattern as $\D$-bound variables
with usage environment matching the linear resources required to type the
corresponding constructor argument in the scrutinee. The resources required to
type each constructor sub-expression come annotated in the judgement (as
${:}_{\ov{\D_i}}$). Unrestricted fields of the constructor are introduced as
unrestricted variables. We note that the typing environment $\D$ always
contains the resources $\ov{\D_i}$ in uses of the alternative judgement.

@@

To satisfy these two constraints, we introduce a $\textit{Split}$ rule that
allows a linear resource $x{:}_1\s$ to be split into $n$ resources at a given
constructor $K$, where $n$ is the number of linear components of the
constructor, and each resource resulting from the split is \emph{tagged} with
$K$ and the positional index of a linear component (i.e. $x{:}_1\s$ can be
split into $\ov{\lctag{x{:}_1\s}{K_i}}^n$).
%
By assigning to each pattern variable a \emph{fragment} of the scrutinee
resources (with a tag), we require the scrutinee resources to be
$\textsc{Split}$ in order to use pattern variables at all.  Moreover, the
remaining tagged resources cannot be used directly, yet need to still be consumed
exactly once. Thus, the choice to consume the tagged resources via the usage
environments of the other pattern variables is forced, ensuring that no
variable for a constructor's linear component can go unused (the tagged environments are disjoint).

For instance, in the term $\lambda x.~\ccase{f~x}{z~\{K~a~b\to
  (a,b)\}}$, where $x$ is a linear variable, the case alternative is
typed with $\irr{x}$,
the case binder $z$ is introduced as $\z[\irr{x}]$, and the pattern variables
are introduced as $\var[a][\lctag{\irr{x}}{K_1}]$ and
$\var[b][\lctag{\irr{x}}{K_2}]$, assuming both components of $K$ are linear.
%
The occurrences of $a$ and $b$ are well-typed because we can first
$\textsc{Split}$ $[x{:}_1\s]$ into $\lctag{[x{:}_1\s]}{K_1}, \lctag{[x{:}_1\s]}{K_2}$,
%
% since for any of the scrutinee
% resources to be used by a linear pattern-bound variable usage, the
% resources must be $Split$ for the fragments corresponding to that
% $\D$-var to be consumed, and, consequently, the remaining fragments
% have to be consumed through the other linear pattern-bound variables.
%
noting that $Split$ can be applied both to relevant and proof irrelevant linear
resources in $\D$.

@@

The implementation of Linear Core as a typechecker does not follow
directly from the description of the type system because Linear Core
is not \emph{syntax-directed}. Specifically, the most challenging
features are splitting linear resources amongst sub-derivations and
consuming fragments of resources through pattern-bound variables.
Instead of non-deterministically splitting linear resources amongst
sub-derivations, we thread input/output linear resources through each
derivation~\cite{DBLP:journals/tcs/CervesatoHP00}.
%
For a scrutinee not in WHNF, pattern variables in a case alternative are
introduced as $\D$-variables with usage environment $\lctag{\irr{\D}}{K_i}$,
where $\D$ are the scrutinee resources and $K_i$ the tag of that pattern
variable. To use these pattern variables, the scrutinee resources must be first
$Split$ into fragments. Instead of guessing which resources need to be split,
the implementation consumes tagged fragments of a resource \emph{as needed},
i.e., only when consuming a resource whose usage environment has a resource
tagged $K_j$ do we $Split$ that resource on $K$ and consume the $K_j$ fragment.
%
Finally, our implementation must infer the usage environments of binders in
a recursive let group before using them to typecheck the let body. We use a
naive $O(n^2)$ algorithm (where $n$ is the number of let bindings) to determine
these usage environments. %Further research regarding
%this inference challenge is discussed in Section~\ref{sec:future-work}.
%
%
%
% We use the following techniques to tackle the \emph{non-syntax-directedness} of
% Linear Core:
%
% \begin{itemize}

>>>>

The alternative judgement $\G;\D \vdash_{alt} \rho \to e :^z_\D \s \Rightarrow
\vp$ is used to type case alternatives, encompassing three
modes that are distinguished by the kind of arrow that is used:
for alternatives of case expressions whose scrutinee is in WHNF
($\Rightarrow_\mathsf{WHNF}$), not in WHNF ($\Rightarrow_\mathsf{NWHNF}$),
and for alternatives agnostic to the WHNF-ness of the scrutinee
($\Rightarrow$), with $\Rightarrow$ also generalizing the other two.
%
% Following the $Case_\textrm{WHNF}$ rule in which we use the $\Mapsto$ alternative
% judgement, t

First, an alternative whose pattern is a constructor with $n > 0$ linear components
under the $\Rightarrow_{\mathsf{WHNF}}$ mode is typed as:
\[
\TypeAltNWHNF
\]
The rule states that, for such a match on a scrutinee already in
WHNF, we introduce the linear components of the pattern as $\D$-bound variables
with usage environment matching the linear resources required to type the
corresponding constructor argument in the scrutinee. The resources required to
type each constructor sub-expression come annotated in the judgement (as
${:}_{\ov{\D_i}}$). Unrestricted fields of the constructor are introduced as
unrestricted variables. We note that the typing environment $\D$ always
contains the resources $\ov{\D_i}$ in uses of the alternative judgement.

@@

To satisfy these two constraints, we introduce a $\textsc{Split}$ rule that
allows a linear resource $x{:}_1\s$ to be split into $n$ resources at a given
constructor $K$, where $n$ is the number of linear components of the
constructor, and each resource resulting from the split is \emph{tagged} with
$K$ and the positional index of a linear component (i.e. $x{:}_1\s$ can be
split into $\ov{\lctag{x{:}_1\s}{K_i}}^n$).
%
By assigning to each pattern variable a \emph{fragment} of the scrutinee
resources (with a tag), we require the scrutinee resources to be
$\textsc{Split}$ in order to use pattern variables at all.  Moreover, the
remaining tagged resources cannot be used directly, yet need to still be consumed
exactly once. Thus, the choice to consume the tagged resources via the usage
environments of the other pattern variables is forced, ensuring that no
variable for a constructor's linear component can go unused (the tagged environments are disjoint).
%
Tagged resources are inspired by fractional permissions in separation logic~\cite{10.5555/1760267.1760273}.

For instance, in the term $\lambda x.~\ccase{f~x}{z~\{K~a~b\to
  (a,b)\}}$, where $x$ is a linear variable, the case alternative is
typed with $\irr{x}$,
the case binder $z$ is introduced as $\z[\irr{x}]$, and the pattern variables
are introduced as $\var[a][\lctag{\irr{x}}{K_1}]$ and
$\var[b][\lctag{\irr{x}}{K_2}]$, assuming both components of $K$ are linear.
%
The occurrences of $a$ and $b$ are well-typed because we can first
$\textsc{Split}$ $[x{:}_1\s]$ into $\lctag{[x{:}_1\s]}{K_1}, \lctag{[x{:}_1\s]}{K_2}$,
%
% since for any of the scrutinee
% resources to be used by a linear pattern-bound variable usage, the
% resources must be $Split$ for the fragments corresponding to that
% $\D$-var to be consumed, and, consequently, the remaining fragments
% have to be consumed through the other linear pattern-bound variables.
%
noting that $\textsc{Split}$ can be applied both to relevant and proof irrelevant linear
resources in $\D$.

@@

The implementation of Linear Core as a typechecker does not follow
directly from the description of the type system because Linear Core
is not \emph{syntax-directed}. Specifically, the most challenging
features are splitting linear resources amongst sub-derivations and
consuming fragments of resources through pattern-bound variables.
Instead of non-deterministically splitting linear resources amongst
sub-derivations, we thread input/output linear resources through each
derivation~\cite{DBLP:journals/tcs/CervesatoHP00}.
%
For a scrutinee not in WHNF, pattern variables in a case alternative are
introduced as $\D$-variables with usage environment $\lctag{\irr{\D}}{K_i}$,
where $\D$ are the scrutinee resources and $K_i$ the tag of that pattern
variable. To use these pattern variables, the scrutinee resources must be first
$\textsc{Split}$ into fragments. Instead of guessing which resources need to be split,
the implementation consumes tagged fragments of a resource \emph{as needed},
i.e., only when consuming a resource whose usage environment has a resource
tagged $K_j$ do we $\textsc{Split}$ that resource on $K$ and consume the $K_j$ fragment.
%
Finally, our implementation must infer the usage environments of binders in
a recursive let group before using them to typecheck the let body. We use a
naive $O(n^2)$ algorithm (where $n$ is the number of let bindings) to determine
these usage environments. %Further research regarding
%this inference challenge is discussed in Section~\ref{sec:future-work}.
%
%
%
% We use the following techniques to tackle the \emph{non-syntax-directedness} of
% Linear Core:
%
% \begin{itemize}
